---
title: "PSY6422-FinalProject"
format: html
editor: visual
---

# Death rates from Alzeimer's disease and other forms of dementia's across Europe from 2000-2021

#### Sophie Ryan

##### 04/12/25

## Background and Data Origins

As a postgraduate Masters student studying cognitive neuroscience and human neuroimaging I am interested in the impact of neurodegenerative diseases on health and populations. This inspired me to create a visualisation looking at how deaths from alzeimers disease and other forms of dementia are increasing or decreasing and how this specifically is impacting health and therefore death rates. The raw data set was obtained from Our World Data website which provided the annual number of deaths from Alzeimer's disease and other forms of dementia per 100,000 people from 2000 to 2021 globally, combining men and women.

A link to the data set can be accessed [Here](https://ourworldindata.org/grapher/death-rate-from-alzheimers-other-dementias-ghe?tab=table)

My visualisation will focus on answering whether there has been an increase in deaths from Alzheimer’s disease and other dementia's across Europe from the year 2000-2021 and whether one specific country has a higher death rate than others due these neurodegenerative diseases.

> **Research Question:** *"How have death rates from Alzeimer's disease and other forms of dementia's changed over time and how has this differed across countries in Europe?"*

## The Project

#### Data Preparation:

Before starting any data preparation the packages needing to be installed and loaded are `tidyverse` and `here`

```{r}
#| include: FALSE
library(here)
library(tidyverse)
```

Loading in the raw data set before I can begin cleaning and organising.

```{r}
#| results: 'hide'
#| message: FALSE
#Loading raw data into R
read_csv(here("death-rate-from-alzheimers-other-dementias-ghe.csv"))

#Reading dataset into a dataframe
df<-read.csv("death-rate-from-alzheimers-other-dementias-ghe.csv")

```

Showing the first 10 rows of the raw data.

```{r}
#| include: FALSE
head(df,10)

```

#### Data Cleaning 

The current raw data includes death rates from Alzheimers and other dementia's across every country in the world. I want to focus just on countries in Europe. Therefore I am going to remove the rows that are irrelevant to my analysis and save a new data set with just the death rates from Alzheimer's and other dementia's across Europe from 2000-2021.

```{r}
#1. Define a vector of European countries to give me something to reduce my raw data based upon 

european_countries<- c(
  "Albania", "Andorra", "Armenia", "Austria", "Azerbaijan", 
  "Belarus", "Belgium", "Bosnia and Herzegovina", "Bulgaria", 
  "Croatia", "Cyprus", "Czechia", "Denmark", "Estonia", "Finland",
  "France", "Georgia", "Germany", "Greece", "Hungary", "Iceland", 
  "Ireland", "Italy", "Kazakhstan", "Latvia", "Lithuania", "Luxembourg", "Malta", "Moldova", "Monaco", "Montenegro", "Netherlands", "North Macedonia", "Norway", "Poland", "Portugal", "Romania", "Russia", "San Marino", "Serbia", "Slovakia", "Slovenia", "Spain", "Sweden", "Switzerland", "Turkey", "Ukraine",
  "United Kingdom" 
)


#2. Filter the data frame 
#Use the command filter to keep all rows where the value in the "Entity" column is also present in the 'european_countries' vector that was created

europe_alzheimers_df<-df %>%
  filter(Entity %in% european_countries)


#3. Displaying first few rows of the new data frame 

print(head(europe_alzheimers_df))


#4. Counting the number of rows in the original date frame and the new data frame to check that the intended rows were removed 

print(paste("Original rows:", nrow(df)))
print(paste("European rows:", nrow(europe_alzheimers_df)))


#5. Saving the new data set with just European countries in it 
write.csv(europe_alzheimers_df, "european_alzeimers_deaths.csv", row.names = FALSE)



```
